Biogeography \& environmental conditions shape bacteriophage \& bacteria interaction networks across the human microbiome

### 2017-10-01 doc/manuscript.tex
Manuscript updates.
Updated manusript with format for new submission.

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Biogeography \& Environmental Conditions Shape Phage \& Bacteria Interaction Networks Across the Human Microbiome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

\title{Biogeography \& Environmental Conditions Shape Phage \& Bacteria
Interaction Networks Across the Human Microbiome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Danai Koutra}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology \& Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Computer Science, University of Michigan, Ann Arbor,
Michigan, 48109}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

\textbf{\emph{Running Title}}: Network Diversity of the Healthy Human
Microbiome

@@

\section{Importance}\label{importance}

@@

Viruses and bacteria are critical components of the human microbiome and
play important roles in health and disease. Bacterial communities have
been associated with disease states, including a range of skin
conditions (1), acute and chronic wound healing conditions (2, 3), and
gastrointestinal diseases, such as inflammatory bowel disease (4, 5),
\emph{Clostridium difficile} infections (6) and colorectal cancer (7,
8). Altered human viromes (virus communities consisting primarily of
bacteriophages) also have been associated with diseases and
perturbations, including inflammatory bowel disease (5, 9), periodontal
disease (10), spread of antibiotic resistance (11), and others (12--17).
Viruses act in concert with their microbial hosts as a single ecological
community (18). Viruses influence their living microbial host
communities through processes including lysis, host gene expression
modulation (19), influence on evolutionary processes such as horizontal
gene transfer (22) or antagonistic co-evolution (26), and alteration of
ecosystem processes and elemental stoichiometry (27).

Previous human microbiome work has focused on bacterial and viral
communities, but have reduced them to two separate communities by
studying them independently (5, 9, 10, 12--17). This approach fails to
capture the complex dynamics of interacting bacteria and phage
communities, which frequently share genetic information and work
together to maintain stable ecosystems. Removal of bacteria or phages
can disrupt or even collapse those ecosystems (18, 28--37).
Relationship-based network approaches allow us to capture this
interaction information. Studying such bacteria-phage interactions
through community-wide networks built from inferred relationships could
offer further insights into the drivers of human microbiome diversity
across body sites and enable the study of human microbiome network
dynamics overall.

In this study, we characterized human-associated bacterial and phage
communities by their inferred relationships using three published paired
virus and bacteria-dominated whole community metagenomic datasets (13,
14, 38, 39). We leveraged machine learning and graph theory techniques
to establish and explore the human bacteria-phage network diversity
therein. This approach built upon previous large-scale phage-bacteria
network analyses by inferring interactions from metagenomic datasets,
rather than culture-dependent data (33), which is limited in the scale
of possible experiments and analyses. Our metagenomic interaction
inference model improved upon previous models of phage-host predictions
that have utilized a variety of techniques, such as linear models to
predict bacteria-phage co-occurrence using taxonomic assignments (40),
and nucleotide similarity models that were applied to both whole virus
genomes (41) and related clusters of whole and partial virus genomes
(42). Our approach uniquely included protein interaction data and was
validated based on experimentally determined positive and negative
interactions (i.e.~who does and does not infect whom). Through this
approach we were able to provide a basic understanding of the network
dynamics associated with phage and bacterial communities on and in the
human body. By building and utilizing a microbiome network, we found
that different people, body sites, and anatomical locations not only
support distinct microbiome membership and diversity (13, 14, 38, 39,
43--45), but also support ecological communities with distinct
communication structures and propensities toward community instability.
Through an improved understanding of network structures across the human
body, we empower future studies to investigate how these communities
dynamics are influenced by disease states and the overall impact they
may have on human health.

@@

We studied the differences in virus-bacteria interaction networks across
healthy human bodies by leveraging previously published shotgun sequence
datasets of purified viral metagenomes (viromes) paired with
bacteria-dominated whole community metagenomes. Our study contained
three datasets that explored the impact of diet on the healthy human gut
virome (14), the impact of anatomical location on the healthy human skin
virome (13), and the viromes of monozygotic twins and their mothers (38,
39). We selected these datasets because their virome samples were
subjected to virus-like particle (VLP) purification, which removed
contaminating DNA from human cells, bacteria, etc. To this end, the
publishing authors employed combinations of filtration, chloroform/DNase
treatment, and cesium chloride gradients to eliminate organismal DNA
(e.g.~bacteria, human, fungi, etc) and thereby allow for direct
assessment of both the extracellular and fully-assembled intracellular
virome \textbf{(Supplemental Figure \ref{SequenceStats} A-B)} (14, 39).
Each research group reported quality control measures to ensure the
purity of the virome sequence datasets, using both computational and
molecular techniques (e.g.~16S rRNA gene qPCR) \textbf{(Table S1)}.
These reports confirmed that the virome libraries consisted of highly
purified virus genomic DNA.

The bacterial and viral sequences from these studies were quality
filtered and assembled into contigs. We further grouped the related
bacterial and phage contigs into operationally defined units based on
their k-mer frequencies and co-abundance patterns, similar to previous
reports \textbf{(Supplemental Figure \ref{ContigStats} -
\ref{ClusterStats})} (42). We referred to these operationally defined
groups of related contigs as operational genomic units (OGUs). Each OGU
represented a genomically similar sub-population of either bacteria or
phages. Contig lengths within clusters ranged between \(10^{3}\) and
\(10^{5.5}\) bp \textbf{(Supplemental Figure \ref{ContigStats} -
\ref{ClusterStats})}.

While supplementing the previous quality control measures \textbf{(Table
S1)} we found that, in light of the rigorous purification and quality
control during sample collection and preparation, 77\% (228 / 298
operational genomic units) still had some nucleotide similarity to a
given bacterial reference genome (e-value \textless{} \(10^{-25}\)). As
absence of bacterial contamination had been confirmed by sensitive
molecular methods \textbf{(Table S1)}, we interpreted this as evidence
that the majority of the gut and skin bacteriophages were temperate and
thereby shared elements with bacterial reference genomes that contained
similar integrated phages when sequenced--a trend previously reported
(14). Additionally, we identified two OGUs as being complete phages
using the stringent Virsorter phage identification algorithm (class 1
confidence group) (47).

The whole metagenomic shotgun sequence samples primarily consisted of
bacteria, with an average viral relative abundance of 0.4\%
\textbf{(Table S1)} (13, 14, 38, 39). We found that only 2\% (6 / 280
OGUs) of bacterial OGUs had significantly strong nucleotide similarity
to phage reference genomes (e-value \textless{} \(10^{-25}\)) (13, 14,
38, 39). No OGUs were confidently identified as lytic or temperate phage
OGUs in the bacterial dataset using the Virsorter algorithm (47).
Together this suggests only minimal ``contamination'' of the bacterial
OGUs.

@@

We predicted which phage OGUs infected which bacterial OGUs using a
random forest model trained on experimentally validated infectious
relationships from six previous publications (41, 48--52). Only bacteria
and phages were used in the model. The training set contained 43 diverse
bacterial species and 30 diverse phage strains, including both broad and
specific ranges of infection \textbf{(Supplemental Figure
\ref{ValidationOverview} A - B)}. While it is true that there are more
known phages that infect bacteria, we were limited by the information
confirming which phages do not infect certain bacteria and attempted to
keep the numbers of positive and negative interactions similar. Phages
with linear and circular genomes, as well as ssDNA and dsDNA genomes,
were included in the analysis. Because we used DNA sequencing studies,
RNA phages were not considered \textbf{(Supplemental Figure
\ref{ValidationOverview} C-D)}. This training set included both positive
relationships (a phage infects a bacterium) and negative relationships
(a phage does not infect a bacterium). This allowed us to validate the
false positive and false negative rates associated with our candidate
models, thereby building upon previous work that only considered
positive relationships (41).

Four phage and bacterial genomic features were used in a random forest
model to predict infectious relationships between bacteria and phages:
1) genome nucleotide similarities, 2) gene amino acid sequence
similarities, 3) bacterial Clustered Regularly Interspaced Short
Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4)
similarity of protein families associated with experimentally identified
protein-protein interactions (53). The resulting random forest model was
assessed using nested cross validation, and the median area under its
receiver operating characteristic (ROC) curve was 0.788, the median
model sensitivity was 0.952, and median specificity was 0.615
\textbf{(Figure \ref{RocCurve} A)}. The most important predictor in the
model was amino acid similarity between genes, followed by nucleotide
similarity of whole genomes \textbf{(Figure \ref{RocCurve} B)}. Protein
family interactions were moderately important to the model, and CRISPRs
were largely uninformative, due to the minimal amount of identifiable
CRISPRs in the dataset and their redundancy with the nucleotide
similarity methods \textbf{(Figure \ref{RocCurve} B)}. Approximately one
third of the training set relationships yielded no score and therefore
were unable to be assigned an interaction prediction \textbf{(Figure
\ref{RocCurve} C)}.

We used our random forest model to classify the relationships between
bacteria and phage operational genomic units, which were then used to
build the interactive network. The master network contained the three
studies as sub-networks, which themselves each contained sub-networks
for each sample \textbf{(Figure \ref{RocCurve} D)}. Metadata including
study, sample ID, disease, and OGU abundance within the community were
stored in the master network for parsing in downstream analyses
\textbf{(Supplemental Figure \ref{NetworkDiagram})}. Bacterial and phage
relative abundance was recorded in each sample for each OGU and the
weight of the edge connecting those OGUs was calculated as a function of
those relative abundance values. The separate extraction of the phage
and bacterial libraries ensured a more accurate measurement of the
microbial communities, as has been outlined previously (54, 55). The
master network was highly connected and contained 72,287 infectious
relationships among 578 nodes, representing 298 phages and 280 bacteria.
Although the network was highly connected, not all relationships were
present in all samples. Relationships were weighted by the relative
abundances of their associated bacteria and phages. Like the master
network, the skin network exhibited a diameter of 4 (measure of graph
size; the greatest number of traversed vertices required between two
vertices) and included 576 (297 phages, 279 bacteria, 99.7\% total) and
72,127 (99.8\%) of the master network nodes and edges, respectively
\textbf{(Figure \ref{RocCurve} E - F)}. The phages and bacteria in the
diet and twin sample sets were more sparsely related, with the diet
study consisting of 89 (41 phages, 48 bacteria) nodes and 5,566
relationships, and the twin study containing 137 (36 phages, 101
bacteria) nodes and 17,250 relationships \textbf{(Figure \ref{RocCurve}
E - F)}.

@@

Diet is a major environmental factor that influences resource
availability and gut microbiome composition and diversity, including
bacteria and phages (14, 56, 57). Previous work in isolated
culture-based systems has suggested that changes in nutrient
availability are associated with altered phage-bacteria network
structures (30), although this has yet to be tested in humans. We
therefore hypothesized that a change in diet would also be associated
with a change in virome-microbiome network structure in the human gut.

We evaluated the diet-associated differences in gut virome-microbiome
network structure by quantifying how central each sample's network was
on average. We accomplished this by utilizing two common centrality
metrics: degree centrality and closeness centrality. Degree centrality,
the simplest centrality metric, was defined as the number of connections
each phage made with each bacterium. We supplemented measurements of
degree centrality with measurements of closeness centrality. Closeness
centrality is a metric of how close each phage or bacterium is to all of
the other phages and bacteria in the network. A higher closeness
centrality suggests that the effects of genetic information or altered
abundance would be more impactful to all other microbes in the system. A
network with higher average closeness centrality also indicates an
overall greater degree of connections, which suggests a greater
resilience against instability. This is because more highly connected
networks are more stable as a result of pathway dependencies and the
unlikelihood that a randomly removed bacteria or phage would cause major
divisions across the network (30, 58). We used this information to
calculate the average connectedness per sample, which was corrected for
the maximum potential degree of connectedness. Unfortunately our dataset
was insufficiently powered to make strong conclusions toward this
hypothesis, but this is an interesting observation that warrants further
investigation.

@@

Skin and gut community membership and diversity are highly personal,
with people remaining more similar to themselves than to other people
over time (13, 59, 60). We therefore hypothesized that this personal
conservation extended to microbiome network structure. We addressed this
hypothesis by calculating the degree of dissimilarity between each
subject's network, based on phage and bacteria abundance and centrality.
We quantified phage and bacteria centrality within each sample graph
using the weighted eigenvector centrality metric. This metric defines
central phages as those that are highly abundant (\(A_{O}\) as defined
in the methods) and infect many distinct bacteria which themselves are
abundant and infected by many other phages. Similarly, bacterial
centrality was defined as those bacteria that were both abundant and
connected to numerous phages that were themselves connected to many
bacteria. We then calculated the similarity of community networks using
the weighted eigenvector centrality of all nodes between all samples.
Samples with similar network structures were interpreted as having
similar capacities for maintaining stability and transmitting genetic
material.

@@

Extensive work has illustrated differences in diversity and composition
of the healthy human skin microbiome between anatomical sites, including
bacteria, virus, and fungal communities (13, 44, 59). These communities
vary by degree of skin moisture, oil, and environmental exposure. As
viruses are known to influence microbial diversity and community
composition, we hypothesized that microbe-virus network structure would
be specific to anatomical sites, as well. To test this, we evaluated the
changes in network structure between anatomical sites within the skin
dataset.

@@

Foundational work has provided a baseline understanding of the human
microbiome by characterizing bacterial and viral diversity across the
human body (13, 14, 43--45, 61). Here, we offer an initial understanding
of how phage-bacteria networks differ throughout the human body, so as
to provide a baseline for future studies of how and why microbiome
networks differ in disease states. We developed and implemented a
network-based analytical model to evaluate the basic properties of the
human microbiome through bacteria and phage relationships, instead of
membership or diversity alone. This enabled the application of network
theory to provide a new perspective on complex ecological communities.
We utilized metrics of connectivity to model the extent to which
communities of bacteria and phages interact through mechanisms such as
horizontal gene transfer, modulated bacterial gene expression, and
alterations in abundance.

Just as gut microbiome and virome composition and diversity are
conserved in individuals (13, 43, 44, 60), gut and skin microbiome
network structures were conserved within individuals over time. Gut
network structure was not conserved among family members. These findings
suggested that the community properties inferred from microbiome
interaction network structures, such as stability (meaning a more highly
connected network is more stable because a randomly removed bacteria or
phage node is less likely to divide or disintegrate (30, 58) the overall
network), the potential for horizontal gene transfer between members,
and co-evolution of populations, were person-specific. These properties
may be impacted by personal factors ranging from the body's immune
system to external environmental conditions, such as climate and diet.

@@

https://github.com/SchlossLab/Hannigan\_ConjunctisViribus\_mSystems\_2017

@@

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Biogeography \& environmental conditions shape bacteriophage \& bacteria interaction networks across the human microbiome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

\title{Biogeography \& environmental conditions shape bacteriophage \& bacteria
interaction networks across the human microbiome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Danai Koutra}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109, USA}
                    \affil[2]{\nextfont Department of Ecology \& Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109, USA}
                    \affil[3]{\nextfont Department of Computer Science, University of Michigan, Ann Arbor,
Michigan, 48109, USA}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

\textbf{\emph{Short Title}}: Network Diversity of the Healthy Human
Microbiome

@@

\section{Author Summary}\label{author-summary}

@@

Viruses and bacteria are critical components of the human microbiome and
play important roles in health and disease. Bacterial communities have
been associated with disease states, including a range of skin
conditions {[}1{]}, acute and chronic wound healing conditions
{[}2,3{]}, and gastrointestinal diseases, such as inflammatory bowel
disease {[}4,5{]}, \emph{Clostridium difficile} infections {[}6{]} and
colorectal cancer {[}7,8{]}. Altered human viromes (virus communities
consisting primarily of bacteriophages) also have been associated with
diseases and perturbations, including inflammatory bowel disease
{[}5,9{]}, periodontal disease {[}10{]}, spread of antibiotic resistance
{[}11{]}, and others {[}12--17{]}. Viruses act in concert with their
microbial hosts as a single ecological community {[}18{]}. Viruses
influence their living microbial host communities through processes
including lysis, host gene expression modulation {[}19{]}, influence on
evolutionary processes such as horizontal gene transfer {[}22{]} or
antagonistic co-evolution {[}26{]}, and alteration of ecosystem
processes and elemental stoichiometry {[}27{]}.

Previous human microbiome work has focused on bacterial and viral
communities, but have reduced them to two separate communities by
studying them independently {[}5,9,10,12--17{]}. This approach fails to
capture the complex dynamics of interacting bacteria and phage
communities, which frequently share genetic information and work
together to maintain stable ecosystems. Removal of bacteria or phages
can disrupt or even collapse those ecosystems {[}18,28--37{]}.
Relationship-based network approaches allow us to capture this
interaction information. Studying such bacteria-phage interactions
through community-wide networks built from inferred relationships could
offer further insights into the drivers of human microbiome diversity
across body sites and enable the study of human microbiome network
dynamics overall.

In this study, we characterized human-associated bacterial and phage
communities by their inferred relationships using three published paired
virus and bacteria-dominated whole community metagenomic datasets
{[}13,14,38,39{]}. We leveraged machine learning and graph theory
techniques to establish and explore the human bacteria-phage network
diversity therein. This approach built upon previous large-scale
phage-bacteria network analyses by inferring interactions from
metagenomic datasets, rather than culture-dependent data {[}33{]}, which
is limited in the scale of possible experiments and analyses. Our
metagenomic interaction inference model improved upon previous models of
phage-host predictions that have utilized a variety of techniques, such
as linear models to predict bacteria-phage co-occurrence using taxonomic
assignments {[}40{]}, and nucleotide similarity models that were applied
to both whole virus genomes {[}41{]} and related clusters of whole and
partial virus genomes {[}42{]}. Our approach uniquely included protein
interaction data and was validated based on experimentally determined
positive and negative interactions (i.e.~who does and does not infect
whom). Through this approach we were able to provide a basic
understanding of the network dynamics associated with phage and
bacterial communities on and in the human body. By building and
utilizing a microbiome network, we found that different people, body
sites, and anatomical locations not only support distinct microbiome
membership and diversity {[}13,14,38,39,43--45{]}, but also support
ecological communities with distinct communication structures and
propensities toward community instability. Through an improved
understanding of network structures across the human body, we empower
future studies to investigate how these communities dynamics are
influenced by disease states and the overall impact they may have on
human health.

@@

We studied the differences in virus-bacteria interaction networks across
healthy human bodies by leveraging previously published shotgun sequence
datasets of purified viral metagenomes (viromes) paired with
bacteria-dominated whole community metagenomes. Our study contained
three datasets that explored the impact of diet on the healthy human gut
virome {[}14{]}, the impact of anatomical location on the healthy human
skin virome {[}13{]}, and the viromes of monozygotic twins and their
mothers {[}38,39{]}. We selected these datasets because their virome
samples were subjected to virus-like particle (VLP) purification, which
removed contaminating DNA from human cells, bacteria, etc. To this end,
the publishing authors employed combinations of filtration,
chloroform/DNase treatment, and cesium chloride gradients to eliminate
organismal DNA (e.g.~bacteria, human, fungi, etc) and thereby allow for
direct assessment of both the extracellular and fully-assembled
intracellular virome \textbf{(Supplemental Figure \ref{SequenceStats}
A-B)} {[}14,39{]}. Each research group reported quality control measures
to ensure the purity of the virome sequence datasets, using both
computational and molecular techniques (e.g.~16S rRNA gene qPCR)
\textbf{(Table S1)}. These reports confirmed that the virome libraries
consisted of highly purified virus genomic DNA.

The bacterial and viral sequences from these studies were quality
filtered and assembled into contigs. We further grouped the related
bacterial and phage contigs into operationally defined units based on
their k-mer frequencies and co-abundance patterns, similar to previous
reports \textbf{(Supplemental Figure \ref{ContigStats} -
\ref{ClusterStats})} {[}42{]}. We referred to these operationally
defined groups of related contigs as operational genomic units (OGUs).
Each OGU represented a genomically similar sub-population of either
bacteria or phages. Contig lengths within clusters ranged between
\(10^{3}\) and \(10^{5.5}\) bp \textbf{(Supplemental Figure
\ref{ContigStats} - \ref{ClusterStats})}.

While supplementing the previous quality control measures \textbf{(Table
S1)} we found that, in light of the rigorous purification and quality
control during sample collection and preparation, 77\% (228 / 298
operational genomic units) still had some nucleotide similarity to a
given bacterial reference genome (e-value \textless{} \(10^{-25}\)). As
absence of bacterial contamination had been confirmed by sensitive
molecular methods \textbf{(Table S1)}, we interpreted this as evidence
that the majority of the gut and skin bacteriophages were temperate and
thereby shared elements with bacterial reference genomes that contained
similar integrated phages when sequenced--a trend previously reported
{[}14{]}. Additionally, we identified two OGUs as being complete phages
using the stringent Virsorter phage identification algorithm (class 1
confidence group) {[}47{]}.

The whole metagenomic shotgun sequence samples primarily consisted of
bacteria, with an average viral relative abundance of 0.4\%
\textbf{(Table S1)} {[}13,14,38,39{]}. We found that only 2\% (6 / 280
OGUs) of bacterial OGUs had significantly strong nucleotide similarity
to phage reference genomes (e-value \textless{} \(10^{-25}\))
{[}13,14,38,39{]}. No OGUs were confidently identified as lytic or
temperate phage OGUs in the bacterial dataset using the Virsorter
algorithm {[}47{]}. Together this suggests only minimal
``contamination'' of the bacterial OGUs.

@@

We predicted which phage OGUs infected which bacterial OGUs using a
random forest model trained on experimentally validated infectious
relationships from six previous publications {[}41,48--52{]}. Only
bacteria and phages were used in the model. The training set contained
43 diverse bacterial species and 30 diverse phage strains, including
both broad and specific ranges of infection \textbf{(Supplemental Figure
\ref{ValidationOverview} A - B)}. While it is true that there are more
known phages that infect bacteria, we were limited by the information
confirming which phages do not infect certain bacteria and attempted to
keep the numbers of positive and negative interactions similar. Phages
with linear and circular genomes, as well as ssDNA and dsDNA genomes,
were included in the analysis. Because we used DNA sequencing studies,
RNA phages were not considered \textbf{(Supplemental Figure
\ref{ValidationOverview} C-D)}. This training set included both positive
relationships (a phage infects a bacterium) and negative relationships
(a phage does not infect a bacterium). This allowed us to validate the
false positive and false negative rates associated with our candidate
models, thereby building upon previous work that only considered
positive relationships {[}41{]}.

Four phage and bacterial genomic features were used in a random forest
model to predict infectious relationships between bacteria and phages:
1) genome nucleotide similarities, 2) gene amino acid sequence
similarities, 3) bacterial Clustered Regularly Interspaced Short
Palindromic Repeat (CRISPR) spacer sequences that target phages, and 4)
similarity of protein families associated with experimentally identified
protein-protein interactions {[}53{]}. The resulting random forest model
was assessed using nested cross validation, and the median area under
its receiver operating characteristic (ROC) curve was 0.788, the median
model sensitivity was 0.952, and median specificity was 0.615
\textbf{(Figure \ref{RocCurve} A)}. The most important predictor in the
model was amino acid similarity between genes, followed by nucleotide
similarity of whole genomes \textbf{(Figure \ref{RocCurve} B)}. Protein
family interactions were moderately important to the model, and CRISPRs
were largely uninformative, due to the minimal amount of identifiable
CRISPRs in the dataset and their redundancy with the nucleotide
similarity methods \textbf{(Figure \ref{RocCurve} B)}. Approximately one
third of the training set relationships yielded no score and therefore
were unable to be assigned an interaction prediction \textbf{(Figure
\ref{RocCurve} C)}.

We used our random forest model to classify the relationships between
bacteria and phage operational genomic units, which were then used to
build the interactive network. The master network contained the three
studies as sub-networks, which themselves each contained sub-networks
for each sample \textbf{(Figure \ref{RocCurve} D)}. Metadata including
study, sample ID, disease, and OGU abundance within the community were
stored in the master network for parsing in downstream analyses
\textbf{(Supplemental Figure \ref{NetworkDiagram})}. Bacterial and phage
relative abundance was recorded in each sample for each OGU and the
weight of the edge connecting those OGUs was calculated as a function of
those relative abundance values. The separate extraction of the phage
and bacterial libraries ensured a more accurate measurement of the
microbial communities, as has been outlined previously {[}54,55{]}. The
master network was highly connected and contained 72,287 infectious
relationships among 578 nodes, representing 298 phages and 280 bacteria.
Although the network was highly connected, not all relationships were
present in all samples. Relationships were weighted by the relative
abundances of their associated bacteria and phages. Like the master
network, the skin network exhibited a diameter of 4 (measure of graph
size; the greatest number of traversed vertices required between two
vertices) and included 576 (297 phages, 279 bacteria, 99.7\% total) and
72,127 (99.8\%) of the master network nodes and edges, respectively
\textbf{(Figure \ref{RocCurve} E - F)}. The phages and bacteria in the
diet and twin sample sets were more sparsely related, with the diet
study consisting of 89 (41 phages, 48 bacteria) nodes and 5,566
relationships, and the twin study containing 137 (36 phages, 101
bacteria) nodes and 17,250 relationships \textbf{(Figure \ref{RocCurve}
E - F)}.

@@

Diet is a major environmental factor that influences resource
availability and gut microbiome composition and diversity, including
bacteria and phages {[}14,56,57{]}. Previous work in isolated
culture-based systems has suggested that changes in nutrient
availability are associated with altered phage-bacteria network
structures {[}30{]}, although this has yet to be tested in humans. We
therefore hypothesized that a change in diet would also be associated
with a change in virome-microbiome network structure in the human gut.

We evaluated the diet-associated differences in gut virome-microbiome
network structure by quantifying how central each sample's network was
on average. We accomplished this by utilizing two common centrality
metrics: degree centrality and closeness centrality. Degree centrality,
the simplest centrality metric, was defined as the number of connections
each phage made with each bacterium. We supplemented measurements of
degree centrality with measurements of closeness centrality. Closeness
centrality is a metric of how close each phage or bacterium is to all of
the other phages and bacteria in the network. A higher closeness
centrality suggests that the effects of genetic information or altered
abundance would be more impactful to all other microbes in the system. A
network with higher average closeness centrality also indicates an
overall greater degree of connections, which suggests a greater
resilience against instability. This is because more highly connected
networks are more stable as a result of pathway dependencies and the
unlikelihood that a randomly removed bacteria or phage would cause major
divisions across the network {[}30,58{]}. We used this information to
calculate the average connectedness per sample, which was corrected for
the maximum potential degree of connectedness. Unfortunately our dataset
was insufficiently powered to make strong conclusions toward this
hypothesis, but this is an interesting observation that warrants further
investigation.

@@

Skin and gut community membership and diversity are highly personal,
with people remaining more similar to themselves than to other people
over time {[}13,59,60{]}. We therefore hypothesized that this personal
conservation extended to microbiome network structure. We addressed this
hypothesis by calculating the degree of dissimilarity between each
subject's network, based on phage and bacteria abundance and centrality.
We quantified phage and bacteria centrality within each sample graph
using the weighted eigenvector centrality metric. This metric defines
central phages as those that are highly abundant (\(A_{O}\) as defined
in the methods) and infect many distinct bacteria which themselves are
abundant and infected by many other phages. Similarly, bacterial
centrality was defined as those bacteria that were both abundant and
connected to numerous phages that were themselves connected to many
bacteria. We then calculated the similarity of community networks using
the weighted eigenvector centrality of all nodes between all samples.
Samples with similar network structures were interpreted as having
similar capacities for maintaining stability and transmitting genetic
material.

@@

Extensive work has illustrated differences in diversity and composition
of the healthy human skin microbiome between anatomical sites, including
bacteria, virus, and fungal communities {[}13,44,59{]}. These
communities vary by degree of skin moisture, oil, and environmental
exposure. As viruses are known to influence microbial diversity and
community composition, we hypothesized that microbe-virus network
structure would be specific to anatomical sites, as well. To test this,
we evaluated the changes in network structure between anatomical sites
within the skin dataset.

@@

Foundational work has provided a baseline understanding of the human
microbiome by characterizing bacterial and viral diversity across the
human body {[}13,14,43--45,61{]}. Here, we offer an initial
understanding of how phage-bacteria networks differ throughout the human
body, so as to provide a baseline for future studies of how and why
microbiome networks differ in disease states. We developed and
implemented a network-based analytical model to evaluate the basic
properties of the human microbiome through bacteria and phage
relationships, instead of membership or diversity alone. This enabled
the application of network theory to provide a new perspective on
complex ecological communities. We utilized metrics of connectivity to
model the extent to which communities of bacteria and phages interact
through mechanisms such as horizontal gene transfer, modulated bacterial
gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are
conserved in individuals {[}13,43,44,60{]}, gut and skin microbiome
network structures were conserved within individuals over time. Gut
network structure was not conserved among family members. These findings
suggested that the community properties inferred from microbiome
interaction network structures, such as stability (meaning a more highly
connected network is more stable because a randomly removed bacteria or
phage node is less likely to divide or disintegrate {[}30,58{]} the
overall network), the potential for horizontal gene transfer between
members, and co-evolution of populations, were person-specific. These
properties may be impacted by personal factors ranging from the body's
immune system to external environmental conditions, such as climate and
diet.

@@

https://github.com/SchlossLab/Hannigan\_ConjunctisViribus\_ploscompbio\_2017

@@

\section{Author Contributions}\label{author-contributions}
